EvoSwarm: A Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-05-15 draft.tex
More updates

\subsection{EvoSwarm from the User Point of View}

\subsubsection{EvoSwarm Configuration}
\label{sec:evoswarm:config}
The administrator role is responsible for
editing the Docker Compose configuration files that reference the images created
by developers. To deploy the application, he or she needs to edit the {\tt
docker-compose.yml} file to specify the type and number of containers required
for the experiments. The {\tt docker-compose} application is responsible for
starting all the containers, and downloading new versions if it is necessary,
by executing the command {\tt docker-compose up}  inside the EvoSwarm directory

% Also, it's a bit short. Maybe merge it into the next? - JJ

% Again, not a role - JJ
\subsubsection{Experiment Execution}
\label{sec:exeperimentalDefinition}

After starting the application, users can push several Experiment Definition 
files at a time and start monitoring the current execution in the terminal.
After an experiment is finished, the user must execute another script
that takes as argument the experiment id to generate a collection of files containing all the
data generated by the experiment, again in a JSON format. An example of an 
experiment definition is shown in Listing \ref{code:exp}. We give more details about the 
data contained in the file in the next section.

\begin{lstlisting}[language=Python, caption=Experiment definition example, label=code:exp]
{
"DIM_CONFIGURATION" : 
 {
 "2":{"GEN":40,"POP_SIZE": 50,"MAX_ITER":10,"MESSAGES":10},
 "3":{"GEN":25,"POP_SIZE": 60,"MAX_ITER":20,"MESSAGES":10},
 "5":{"GEN":28,"POP_SIZE": 60,"MAX_ITER":30,"MESSAGES":10},
 "10":{"GEN":50,"POP_SIZE": 70,"MAX_ITER":30,"MESSAGES":10},
 "20":{"GEN":66,"POP_SIZE":100,"MAX_ITER":30,"MESSAGES":10},
 "40":{"GEN":80,"POP_SIZE":125,"MAX_ITER":40,"MESSAGES":10}
 },
"GA_WORKER_RATIO" : 0.5,
"FUNCTIONS": [4],
"DIMENSIONS" : [2, 3, 5, 10, 20, 40],
"CXPB_RND": [0.2, 0.6],
"MUTPB_RND": [0.1, 0.3]
}
\end{lstlisting}

% This is totally unexpected. Whose role is this? You didn't even introduce it. Also, extremely short - JJ
% This is one of the applications described in the Workflow Figure. But this is better explained in other places.
% We can delete it (even from the Figure)  
% Will leave it until we decide - M 
\subsubsection{Using the COCO Framework}

Researchers can use these files to analyze and plot experiment data.  Python
scripts are included in the repository to plot the running times for the
experiments, and all other plots used in this paper;  there is also a script to
generate the files needed by the COCO framework \cite{hansen2016coco}, which can generate standard
comparisons against other methods.

@@

\subsection{Deploying to a Cloud Provider}
\label{cloud-aws}
Container-based applications describe the deployment of loosely coupled
processes in a consistent and portable way, in every phase of the life cycle of
an application, from development through testing to production independently on
the kind of infrastructure. The term cloud-native describes these
container-based environments because they are deployed on an infrastructure that
abstracts the underlying compute, storage, and networking primitives. This
infrastructure does need to be a cloud provider; it could be a developer's
laptop, a single server, or a production environment. When we deploy a
cloud-native application to a cloud provider, we have some more options to
choose:

@@

As a proof-of-concept, we chose the third option, that is, a container orchestration service, and deployed the EvoSwarm application in Amazon Elastic
Container Service (ECS),  since it needs fewer changes in configuration 
(from deploying it locally using {\tt docker-compose}) and it is has a lower cost than the Kubernetes alternative. % Remember to mention cost at the end - JJ

@@

We deployed two configurations, one that included all containers in a single task,
and another one, where we assigned each worker to a different task and the controller, 
redis, and setup in another, for a total of nine concurrent Fargate tasks.

@@

@@

% Write some introduction to this subsubsection. 
The results are shown in Figure~\ref{fig:aws-results} and Table~\ref{tab:aws}.

As we can see, the local environment (see section \ref{setup}) achieved a higher
ratio of evaluations per second (in bold). Still, we must consider that there is a minimum
network overhead in a local workstation and a clock speed of 3.8 GHz. The first
option (labeled Single) is straightforward to deploy, needing only a single
{\tt docker-compose} file, and one command to execute. But with the limit of 4CPUs and
ten containers, it is the slowest option. The second option (labeled Multi), is
faster, and not too far from the local execution, especially when solving the
$f_4$ in 20 dimensions (underline in table). With the capacity for increasing the number of worker
containers, it is reasonable to think that we can achieve a higher number of
evaluations with an appropriate configuration. Nevertheless, the goal of this
experiment was not to increase the number of evaluations but to test the
flexibility of a cloud-native application.
% Mention costs - JJ

@@

% Emphasize these results for the purpose of the paper - JJ

% You should clarify that, before, you were explaining how to set up
% an experiment. In this section, you are explaining the experiments
% that you are doing in this paper. But you should relate both
% sections to the general objectives of the paper. In the prior
% section, you want to prove that EvoSwarm can be used by the public
% to create reproductive research (and maybe you should add a small
% conclusion sub or sub-sub section), in the next you explain the
% experiment we are doing for this paper. It would be interesting to
% make references from the next section to the previous one,
% indicating how the general setup of EvoSwarm helps _us_ to create
% this experiment - JJ

@@

We aimed to learn if the proposed solution can efficiently
improve the scalability and performance of population-based optimization
algorithms. % If you want to prove it *improves*, you need to provide a baseline. 
            %What will be your baseline? You need to specify anyway 
            %in which way it really improves - JJ
            % The baseline is a single population GA or PSO? - M
% For instance. But I was actually looking for another word that does not imply a comparison - JJ
Hence, in the following sections, we verify the following:

\begin{itemize}
\item Can we improve the execution time of the algorithm by adding populations
 and serverless functions to the system and, in particular, what is its effect on the scalability of the system? % Improve with respect to what? Single population? Please explain carefully the issues and how scaling is related to the number of populations. - JJ

\item Does having a multi-population enabled platform, with heterogeneous populations and
the support for mixing search strategies,  increase the performance of the
search by needing fewer function evaluations than a homogeneous setting?

% Moved reproducibility question from here - M
\end{itemize}

% Also, point forward to the section (or subsections) that will be devoted to proving each point.
% BTW, they should probably be subsections.

\subsection{General Experimental Setup} 
\label{sec:general-setup} % To be renamed

@@

The experiments can be easily reproduced by using the configuration
files the EvoSwarm application uses to run the noiseless BBOB testbed
experiments as seen in Section~\ref{sec:exeperimentalDefinition}. 
As a first step, we must deploy the docker application using a
{\tt docker-compose.yml} file, as described in Section~\ref{sec:evoswarm:config}. 
Then a JSON file containing the configuration parameters of the
experiment has to be provided.  Table \ref{tab:params} shows an example of these
parameters. The {\em GA-PSO Ratio} parameter indicates the proportion of
populations that will use the GA algorithm.  In the example, with a value of
$0.50$ there will be about the same proportion of GA and PSO populations. If we
specify a value of $0$, this will give us an algorithm with only PSO
populations, and finally, a value of  $1$ means that all populations will run
the GA algorithm.

@@

For local experiments we have deployed the EvoSwarm application in a
Desktop PC with an AMD Ryzen 9 3900x 12-core processor with 24 threads with a 3.8 GHz base clock, 
and 16 GB RAM. We used Docker version 19.03.3, build a872fc2f86, and {\tt docker-compose} version 1.21.0, 
in Ubuntu Linux 18.04; coce is run using Python 3.7.5. Container images and the {\tt docker-compose} file are available at
\url{https://hub.docker.com/u/mariosky}, and \url{https://github.com/mariosky/EvoSwarm} respectively. 
The BOBB experiments were performed with COCO \cite{hansen2016coco} version bbob.v15.03 in Python, 
the plots were produced with version 2.3.2.

% Maybe subsection? Also say the question you're answering - JJ
\subsection{Measuring Scalability}
\label{sec:exp1}
% Experiment 1
As we mentioned before % say where - JJ
, reactive systems can scale by adding additional copies
of serverless functions. In our case, we can start additional worker containers
to have the same effect. Other authors, like Salza et al. \cite{salza2019speed},
have used a
similar architecture, which implies that worker nodes need to have at least a certain level
of complexity, in terms of execution time,  to effectively scale on multiple
nodes, tending to linear scalability. In that paper, Salza et al. tested scalability
by simulating the work of nodes by using a sleep function and messages of different sizes.
In a population-based algorithm like EvoSwarm, several parameters can increase or 
decrease the workload of workers:

@@

To evaluate the effect of these parameters on the scalability of the system, we
propose  an experiment in which we selected $f_4$ (Skew Rastrigin-Bueche separable) from the
BBOB testbed. This function has been used because it is computationally demanding,
and in higher dimensions has been difficult for PSO \cite{el2009black} and GAs
\cite{nicolau2009application} to solve with the required FEs. As we
are only comparing in terms of execution time, the results from a single function
can be better understood, since there are fewer factors involved.
Following the procedure described in the previous section, we executed 
two sets of experiments with five and ten initial populations,
repeating each experiment 30 times using  1, 2, 4, and 8 workers.  We kept the same 
maximum number of FEs, changing the relevant parameters for this. 
See Tables \ref{tab:params:10} and \ref{tab:params:5} 
for the complete list. 
%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{time_diff}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes.  }
  % Which function? - JJ
  \label{fig:spworker}
\end{figure}
%
\begin{table}[]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations} % time or rate? - JJ
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrr}
    \hline
              &         & 5 Populations    & 10 Populations \\
     Dimension& Workers & Median of rate & Median of rate  \\
    \hline
          2   & 1w      & 2.96   & 3.15   \\
              & 2w      & 1.84   & 1.72   \\
              & 4w      & 1.22   & \textbf{0.96}   \\
              & 8w      & 1.05   & \textbf{0.59}   \\
    \hline
          3   & 1w      & 6.53   & \textbf{7.76}   \\
              & 2w      & 3.64   & 3.83   \\
              & 4w      & 2.33   & 2.26   \\
              & 8w      & 2.17   &  \textbf{1.36}   \\
    \hline
          5   & 1w      & \textbf{12.33}  & 14.19  \\
              & 2w      & 6.63   & \textbf{7.41}   \\
              & 4w      & 4.25   & 4.34   \\
              & 8w      & 3.67   &  \textbf{2.41}   \\
    \hline
          10  & 1w      & \textbf{31.50}  & 38.30  \\
              & 2w      & \textbf{16.07}  & 19.31  \\
              & 4w      &  9.99   & 10.35  \\
              & 8w      & 8.97   &  \textbf{5.92}  \\
    \hline
          20  & 1w      &  116.35 & 124.51 \\
              & 2w      & \textbf{51.18}  & 63.68  \\
              & 4w      & 32.26  & 33.07  \\
              & 8w      & 28.64  & \textbf{18.25}  \\
    \hline
          40  & 1w      & \textbf{369.89} & 448.89 \\
              & 2w      & \textbf{198.43} & 234.50 \\
              & 4w      & 122.39 &  120.62 \\
              & 8w      & 122.31 & \textbf{63.35} \\
    \hline
  \end{tabular}
\end{table}

@@

But this is a rule of thumb, and it will depend on the degree of heterogeneity,
as well as on the algorithm itself. Some level of heterogeneity can be
implemented by just changing the configuration parameters of each population,
as in the previous experiment, but in this case, we are interested in a heterogeneous search strategy.

>>>>

\subsection{EvoSwarm from the User Point of View} 
% All in a single section - Mario
%\subsubsection{Experiment Configuration and Execution}
\label{sec:evoswarm:config}

The user role is responsible for
editing the {\tt docker-compose} configuration files that reference the images created
by developers. To deploy the application, he or she needs to edit the {\tt
docker-compose.yml} file to specify the type and number of containers required
for the experiments. Then, the {\tt docker-compose} application is responsible for
starting all the containers, and downloading new versions if it is necessary,
by executing the command {\tt docker-compose up}  inside the EvoSwarm directory.

After starting the application, users can push several Experiment Definition 
files at a time and start monitoring the current execution in the terminal.
After an experiment is finished, the user must execute another script
that takes as argument the experiment id to generate a collection of files containing all the
data generated by the experiment, again in a JSON format. An example of an 
experiment definition is shown in Listing \ref{code:exp}. We give more details about the 
data contained in the file in the next section. Users can use these files 
to analyze and plot experiment data.  Python
scripts are included in the repository to plot the running times for the
experiments, and all other plots used in this paper;  there is also a script to
generate the files needed by the COCO framework \cite{hansen2016coco}, which can generate standard
comparisons against other methods.

\begin{lstlisting}[language=Python, caption=Experiment definition example, label=code:exp]
{
"DIM_CONFIGURATION" : 
 {
 "10":{"GEN":50,"POP_SIZE": 70,"MAX_ITER":30,"MESSAGES":10},
 "20":{"GEN":66,"POP_SIZE":100,"MAX_ITER":30,"MESSAGES":10},
 },
"GA_WORKER_RATIO" : 0.5,
"FUNCTIONS": [4],
"DIMENSIONS" : [10, 20],
"CXPB_RND": [0.2, 0.6],
"MUTPB_RND": [0.1, 0.3]
}
\end{lstlisting}

@@

\subsection{Deploying to a Cloud Provider}
\label{cloud-aws}

As a proof-of-concept, we present a brief experiment that is deployed in three
distinct environments, taking advantage that as a container-based application,
EvoSwarm describes the deployment of loosely coupled processes in a consistent
and portable way.

The term cloud-native describes these container-based
environments because we can deploy them on an infrastructure that abstracts the
underlying compute, storage, and networking primitives. This infrastructure does not
need to be a cloud provider; it could be a developer's laptop, a single server,
or a production environment. When we deploy a cloud-native application to a
cloud provider, we have some more options to choose:

%Container-based applications describe the deployment of loosely coupled
%processes in a consistent and portable way, in every phase of the life cycle of
%an application, from development through testing to production independently on
%the kind of infrastructure. The term cloud-native describes these
%container-based environments because they are deployed on an infrastructure that
%abstracts the underlying compute, storage, and networking primitives. This
%infrastructure does need to be a cloud provider; it could be a developer's
%laptop, a single server, or a production environment. When we deploy a
%cloud-native application to a cloud provider, we have some more options to
%choose:

@@

\subsubsection{Setup}
\label{setup:local-cloud}

We chose the third option, that is, a container orchestration service, and
deployed the EvoSwarm application in Amazon Elastic Container Service (ECS),
since it needs fewer changes in configuration (from deploying it locally using
{\tt docker-compose}) and it is has a lower cost than the Kubernetes
alternative.

% Remember to mention cost at the end - JJ

@@

We deployed two ECS configurations, one that included all containers in a single task,
and another one, where we assigned each worker to a different task and the controller, 
redis, and setup in another, for a total of nine concurrent Fargate tasks.

@@

For local experiments, we have deployed the EvoSwarm application in a
Desktop PC with an AMD Ryzen 9 3900x 12-core processor with 24 threads with a 3.8 GHz base clock, 
and 16 GB RAM. We used Docker version 19.03.3, build a872fc2f86, and {\tt docker-compose} version 1.21.0, 
in Ubuntu Linux 18.04; code is run using Python 3.7.5. Container images and the {\tt docker-compose} file are available at
\url{https://hub.docker.com/u/mariosky}, and \url{https://github.com/mariosky/EvoSwarm} respectively.

We reproduced the experiment of Listing~\ref{code:exp} in each environment. The experiment consists of
solving $f_4$ for 10 and 20 dimensions. We give more details about additional parameters of the configuration in Section~\ref{setup}.

@@

% Write some introduction to this subsubsection.

In this experiment, we wanted to test several deployment options for the 
cloud-native application, and have an idea of the computing capacity of each option
measured as the number of function evaluations per second.
The results are shown in Figure~\ref{fig:aws-results} and Table~\ref{tab:aws}. 
As we can see, the local environment (see section \ref{setup}) achieved a higher
ratio of evaluations per second (in bold). Still, we must consider that there is a minimum
network overhead in a local workstation and a clock speed of 3.8 GHz. The first
option (labeled Single) is straightforward to deploy, needing only a single
{\tt docker-compose} file, and one command to execute. But with the limit of 4CPUs and
ten containers, it is the slowest option. The second option (labeled Multi), is
faster, and not too far from the local execution, especially when solving the
$f_4$ in 20 dimensions (underline in table). With the capacity for increasing the number of worker
containers, it is reasonable to think that we can achieve a higher number of
evaluations with an appropriate configuration. Nevertheless, the goal of this
experiment was not to increase the number of evaluations but to test the
flexibility of a cloud-native application. The total cost of testing, configuration, 
and execution of the experiments in ECS was less than \$1.00 US Dollar.

% Mention costs - JJ

@@

The same workflow and configuration files used by this brief experiment, will be
used on the next section, to carry a more extensive evaluation of the framework.
Again, each configuration describe the deployment of containers in a consistent
and portable way, an it was used in each phase of the life cycle, from
development through testing to production independently on the kind of
infrastructure.

% Emphasize these results for the purpose of the paper - JJ
% done - Mario

@@

% You should clarify that, before, you were explaining how to set up
% an experiment. In this section, you are explaining the experiments
% that you are doing in this paper. But you should relate both
% sections to the general objectives of the paper. In the prior
% section, you want to prove that EvoSwarm can be used by the public
% to create reproductive research (and maybe you should add a small
% conclusion sub or sub-sub section), in the next you explain the
% experiment we are doing for this paper. It would be interesting to
% make references from the next section to the previous one,
% indicating how the general setup of EvoSwarm helps _us_ to create
% this experiment - JJ  
% Connected below - M

In the prior section, we wanted to prove that EvoSwarm can be used by 
the public to create reproductive research, but in this section are
using the same framework to learn if the proposed solution can efficiently
improve the scalability and performance of population-based optimization
algorithms. % If you want to prove it *improves*, you need to provide a baseline. 
            %What will be your baseline? You need to specify anyway 
            %in which way it really improves - JJ
            % The baseline is a single population GA or PSO? - M
% For instance. But I was actually looking for another word that does not imply a comparison - JJ
Hence, in the following sections, we verify the following questions:

\begin{itemize}
\item Can we improve the execution time of the algorithm by adding populations
 and serverless functions (workers)  to the system and, in particular,
 what is its effect on the scalability of the system? % Improve with respect to what? Single population? 
                         %Please explain carefully the issues and how scaling is related to the number of populations. - JJ
                         % The question could be that, How scaling is related to number of populations and workers in the system ? - M

\item Does having a multi-population enabled platform, with heterogeneous populations and
the support for mixing search strategies,  increase the performance of the
search by needing fewer function evaluations than a homogeneous setting? 
% Moved reproducibility question from here - M

\end{itemize}

Section~\label{sec:exp1} is devoted to answering the first question, and Section~\label{sec:exp2} to the second.
Next, we describe the general experimental setup to both experiments.

% Also, point forward to the section (or subsections) that will be devoted to proving each point.
% BTW, they should probably be subsections.
% Done - M

\subsection{General Experimental Setup} 
\label{sec:general-setup}

@@

The experiments can be easily reproduced by using the configuration
files the EvoSwarm application uses to run the noiseless BBOB testbed
experiments as seen in Section~\ref{sec:evoswarm:config}. 
As a first step, we must deploy the docker application using a
{\tt docker-compose.yml} file, as described in Section~\ref{sec:evoswarm:config}. 
Then a JSON file containing the configuration parameters of the
experiment has to be provided.  Table \ref{tab:params} shows an example of these
parameters. The {\em GA-PSO Ratio} parameter indicates the proportion of
populations that will use the GA algorithm.  In the example, with a value of
$0.50$ there will be about the same proportion of GA and PSO populations. If we
specify a value of $0$, this will give us an algorithm with only PSO
populations, and finally, a value of  $1$ means that all populations will run
the GA algorithm.

@@

We have deployed the EvoSwarm application locally in the same hardware and
software described in Section~\ref{setup:local-cloud}, additionally the BOBB
experiments were performed with COCO \cite{hansen2016coco} version bbob.v15.03
in Python, the plots were produced with version 2.3.2.

% Maybe subsection? Also say the question you're answering - JJ
% Done, I added the research question at the beginning of the setup question - M
\subsection{Measuring Scalability}
\label{sec:exp1}
% Experiment 1

As we mentioned before in Section~\ref{method}, reactive systems can scale by adding additional copies
of serverless functions. In our case, we can start additional worker containers
to have the same effect. Other authors, like Salza et al. \cite{salza2019speed},
have used a similar architecture, which implies that worker nodes need to have at least a certain level
of complexity, in terms of execution time,  to effectively scale on multiple
nodes, tending to linear scalability. In that paper, Salza et al. tested scalability
by simulating the work of nodes by using a sleep function and messages of different sizes.
In a population-based algorithm like EvoSwarm, several parameters can increase or 
decrease the workload of workers:

@@

To answer the first research question, we need to evaluate the effect of these parameters on the scalability of the system,   we
propose  an experiment in which we selected $f_4$ (Skew Rastrigin-Bueche separable) from the
BBOB testbed. This function has been used because it is computationally demanding,
and in higher dimensions has been difficult for PSO \cite{el2009black} and GAs
\cite{nicolau2009application} to solve with the required FEs. As we
are only comparing in terms of execution time, the results from a single function
can be better understood, since there are fewer factors involved.
Following the procedure described in the previous section, we executed 
two sets of experiments with five and ten initial populations,
repeating each experiment 30 times using  1, 2, 4, and 8 workers.  We kept the same 
maximum number of FEs, changing the relevant parameters for this. 
See Tables \ref{tab:params:10} and \ref{tab:params:5} 
for the complete list. 
%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{time_diff}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes.  }
  \label{fig:spworker}
\end{figure}
%
\begin{table}[]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations} % time or rate? - JJ
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrr}
    \hline
              &         & 5 Populations    & 10 Populations \\
     Dimension& Workers & Median of rate & Median of rate  \\
    \hline
          2   & 1w      & 2.96   & 3.15   \\
              & 2w      & 1.84   & 1.72   \\
              & 4w      & 1.22   & \textbf{0.96}   \\
              & 8w      & 1.05   & \textbf{0.59}   \\
    \hline
          3   & 1w      & 6.53   & \textbf{7.76}   \\
              & 2w      & 3.64   & 3.83   \\
              & 4w      & 2.33   & 2.26   \\
              & 8w      & 2.17   &  \textbf{1.36}   \\
    \hline
          5   & 1w      & \textbf{12.33}  & 14.19  \\
              & 2w      & 6.63   & \textbf{7.41}   \\
              & 4w      & 4.25   & 4.34   \\
              & 8w      & 3.67   &  \textbf{2.41}   \\
    \hline
          10  & 1w      & \textbf{31.50}  & 38.30  \\
              & 2w      & \textbf{16.07}  & 19.31  \\
              & 4w      &  9.99   & 10.35  \\
              & 8w      & 8.97   &  \textbf{5.92}  \\
    \hline
          20  & 1w      &  116.35 & 124.51 \\
              & 2w      & \textbf{51.18}  & 63.68  \\
              & 4w      & 32.26  & 33.07  \\
              & 8w      & 28.64  & \textbf{18.25}  \\
    \hline
          40  & 1w      & \textbf{369.89} & 448.89 \\
              & 2w      & \textbf{198.43} & 234.50 \\
              & 4w      & 122.39 &  120.62 \\
              & 8w      & 122.31 & \textbf{63.35} \\
    \hline
  \end{tabular}
\end{table}

@@

But this is a rule of thumb, and it will depend on the degree of heterogeneity,
as well as on the algorithm itself. Some level of heterogeneity can be
implemented by just changing the configuration parameters of each population,
as in the previous experiment. But in this case, we are interested in 
answering the second research question, and we want to validate if by using 
heterogeneous populations and mixing search strategies,  we can increase the performance of the
search.
